### 2025-04-08 Paper.tex
order refined

%==============================================
%================ Preprocessing ================
%==============================================
\section*{Preprocessing}

The initial phase of our project focused on preprocessing the audio data, which is a crucial step that encompasses several subtasks. The preprocessing aims at preparing the raw material for further analysis.

\subsubsection*{Data Acquisition}
The first step before we could start with the preprocessing involved acquiring the songs that we wanted to use for our project. We wanted to ensure a diverse and representative dataset; thus, we searched for publicly available audio collections of Linkin Park. As we did not find a complete collection, we sourced the audios from YouTube. In total, we curated a corpus consisting of eight albums, each containing an average of 12 tracks. This collection formed the foundation for our subsequent processing and analysis steps.

@@

\subsubsection*{Segmentation}
After the vocal extraction, the next step was to segment the data. Segmenting the audios into smaller chunks is important as it can significantly enhance the model’s performance and the reliability of the forced alignment that followed this step \cite{Rybach2009}. The quality of the alignment heavily depends on the segmentation, as a wrong segmentation will lead to a wrong alignment. The effectiveness of our model is, therefore, closely tied to the goodness of the audio segmentation.

@@

@@

\subsubsection*{Alignment Tools}
We initially used the Montreal Forced Aligner (MFA), which is a widely recognized tool for automatic speech-text alignment. The MFA is built upon the Kaldi speech recognition toolkit. It is designed to provide robust phonetic segmentation by leveraging triphone acoustic models and speaker adaptation techniques \cite{McAuliffe2017}. MFA has been successfully applied to various languages and datasets; however, in our case, the MFA did not reveal convincing results. We attributed this to the fact that the MFA is primarily designed for spoken language, while singing poses some additional challenges, such as extended phonemes and pitch variations.

@@

\subsubsection*{Data Conversion with nnsvs-db-converter}
We used the nnsvs-db-converter, a Python tool for converting vocal databases with HTS mono labels into DiffSinger format to make them compatible with variance models for predicting pitch and timing \cite{nnsvsdbconverter}. It handles sample segmentation by ensuring that each sample has a maximum length and an acceptable number of pauses between segments, and converts silent phonemes (e.g., sil, pau) to standardized representations (SP and AP for breaths). The script requires minimal external dependencies, making it easily accessible. In addition, it supports language definitions that specify vowels and liquids for better segmentation and timing handling, and it can estimate MIDI pitch and timing information for the use of variance models.

\subsubsection*{SlurCutter Exploration}
We also had a look into SlurCutter, which is a specialized tool that is designed to assist in the creation of datasets for DiffSinger. Its primary function is to edit MIDI sequences that are stored in .ds files to facilitate the preparation of data for variance model training \cite{openvpi}. However, it requires a lot of time and effort as this process is done manually, so we decided not to do it on a large scale.

\subsubsection*{Manual Segmentation}
To include more data and improve model performance, we additionally included two live albums in our dataset. As many segments contained screaming from the audience, we went through them and cut out unwanted parts manually with Praat, a tool for speech analysis and phonetic research.

\subsubsection*{Manual Alignment}
In addition to our segmented audio snippets, we fully labeled two complete songs — Over Each Other and The Emptiness Machine — using LabelMakr, which we later used during inference. Since these full-song annotations were created after the short segments had already been labeled, the initial alignments were expectedly poor. To address this, we manually refined the boundaries: we shifted misaligned ones, added missing boundaries, and removed others — particularly SP (speech pause) and AP (aspiration pause) annotations, which frequently appeared in inappropriate positions.

@@

\subsection*{Refined Preprocessing}
As our initial training revealed noisy outputs, we decided to improve our dataset to enhance model performance. We began by manually manipulating two albums, where we removed segments that contained excessive noise, overlapping voices, or rap sections. The result was a cleaned subset of data, which we stored as our clear training data.

@@

%==============================================
%================ Training =====================
%==============================================
\section*{Training}

\section*{Manual Augmentation}

@@

Model 7 (manual + sorted clean)

@@

Model 8

@@

Model 9

@@

%==============================================
%================ Evaluation ===================
%==============================================
\section*{Evaluation}

%==============================================
%============= Mixing and Mastering ============
%==============================================
\section*{Mixing and Mastering}

%==============================================
%================ Conclusion ===================
%==============================================
\section*{Conclusion}

@@

Despite the many different experiments, the results of the models are relatively comparable. While the data augmented models seem to outperform the small and large model, and the fine-tuned vocoder also improved model performance, the final results are still disappointing in quality. Some segments are processed well by (almost) all models,however, the output of other segments is inadequate, as the voice sounds distorted and rough or other background noises appear to be involved. In many segments, the phonemes or phoneme boundaries also appear to be incorrect. This is probably due to the alignment. To address this issue, aligning the data manually would almost certainly lead to a better model as many boundaries are inadequate. However, this step requires a non-negligible amount of time, offering potential for future research. There can also be made improvements within the data augmentation as the project has highlighted the importance of careful and balanced augmentation to avoid degrading the model’s performance.

>>>>

%==============================================
%================ Preprocessing ================
%==============================================
\section{Preprocessing}

\subsection{Initial Preprocessing}
The initial phase of our project focused on preprocessing the audio data, which is a crucial step that encompasses several subtasks. The preprocessing aims at preparing the raw material for further analysis.

\subsubsection{Data Acquisition}
The first step before we could start with the preprocessing involved acquiring the songs that we wanted to use for our project. We wanted to ensure a diverse and representative dataset; thus, we searched for publicly available audio collections of Linkin Park. As we did not find a complete collection, we sourced the audios from YouTube. In total, we curated a corpus consisting of eight albums, each containing an average of 12 tracks. This collection formed the foundation for our subsequent processing and analysis steps.

@@

\subsubsection{Segmentation}
After the vocal extraction, the next step was to segment the data. Segmenting the audios into smaller chunks is important as it can significantly enhance the model’s performance and the reliability of the forced alignment that followed this step \cite{Rybach2009}. The quality of the alignment heavily depends on the segmentation, as a wrong segmentation will lead to a wrong alignment. The effectiveness of our model is, therefore, closely tied to the goodness of the audio segmentation.

@@

\subsubsection{Manual Segmentation}
To include more data and improve model performance, we additionally included two live albums in our dataset. As many segments contained screaming from the audience, we went through them and cut out unwanted parts manually with Praat, a tool for speech analysis and phonetic research.

@@

\subsubsection{Alignment Tools}
We initially used the Montreal Forced Aligner (MFA), which is a widely recognized tool for automatic speech-text alignment. The MFA is built upon the Kaldi speech recognition toolkit. It is designed to provide robust phonetic segmentation by leveraging triphone acoustic models and speaker adaptation techniques \cite{McAuliffe2017}. MFA has been successfully applied to various languages and datasets; however, in our case, the MFA did not reveal convincing results. We attributed this to the fact that the MFA is primarily designed for spoken language, while singing poses some additional challenges, such as extended phonemes and pitch variations.

@@

\subsubsection{Manual Alignment}
In addition to our segmented audio snippets, we fully labeled two complete songs — Over Each Other and The Emptiness Machine — using LabelMakr, which we later used during inference. Since these full-song annotations were created after the short segments had already been labeled, the initial alignments were expectedly poor. To address this, we manually refined the boundaries: we shifted misaligned ones, added missing boundaries, and removed others — particularly SP (speech pause) and AP (aspiration pause) annotations, which frequently appeared in inappropriate positions.

@@

\subsubsection{Data Conversion with nnsvs-db-converter}
We used the nnsvs-db-converter, a Python tool for converting vocal databases with HTS mono labels into DiffSinger format to make them compatible with variance models for predicting pitch and timing \cite{nnsvsdbconverter}. It handles sample segmentation by ensuring that each sample has a maximum length and an acceptable number of pauses between segments, and converts silent phonemes (e.g., sil, pau) to standardized representations (SP and AP for breaths). The script requires minimal external dependencies, making it easily accessible. In addition, it supports language definitions that specify vowels and liquids for better segmentation and timing handling, and it can estimate MIDI pitch and timing information for the use of variance models.

\subsubsection{SlurCutter Exploration}
We also had a look into SlurCutter, which is a specialized tool that is designed to assist in the creation of datasets for DiffSinger. Its primary function is to edit MIDI sequences that are stored in .ds files to facilitate the preparation of data for variance model training \cite{openvpi}. However, it requires a lot of time and effort as this process is done manually, so we decided not to do it on a large scale.

\subsection{Refined Preprocessing}
As our initial training revealed noisy outputs, we decided to improve our dataset to enhance model performance. We began by manually manipulating two albums, where we removed segments that contained excessive noise, overlapping voices, or rap sections. The result was a cleaned subset of data, which we stored as our clear training data.

@@

%==============================================
%================ Training =====================
%==============================================
\section{Training}

\subsection{Manual Augmentation}

@@

\subsubsection{Model 7 (manual + sorted clean)}

@@

\subsubsection{Model 8 (manual + sorted + augmented clean)}

@@

\subsubsection{Model 9 (large + augmented)}

@@

%==============================================
%================ Evaluation ===================
%==============================================
\section{Evaluation}

%==============================================
%============= Mixing and Mastering ============
%==============================================
\section{Mixing and Mastering}

%==============================================
%================ Conclusion ===================
%==============================================
\section{Conclusion}

@@

Despite the many different experiments, the results of the models are relatively comparable. Some segments are processed well by (almost) all models,however, the output of other segments is inadequate, as the voice sounds distorted and rough or other background noises appear to be involved. In many segments, the phonemes or phoneme boundaries also appear to be incorrect. To quantify model performance, we applied two types of evaluation: predicted Mean Opinion Score (MOS) using SingMOS and a mel-spectrogram histogram comparison. The variance+vocoder model consistently outperformed others across both evaluation types, with MOS scores approaching the ground truth and strong spectral similarity to real vocals. However, even this best-performing model could not fully replicate the expressive detail and clarity of the original vocals, often struggling with distorted phonemes, rough transitions, or alignment errors.

A probable issue could be the automatic alignment. To address this problem, aligning the data manually would almost certainly lead to a better model as many boundaries are inadequate. However, this step requires a non-negligible amount of time, offering potential for future research. There can also be made improvements within the data augmentation as the project has highlighted the importance of careful and balanced augmentation to avoid degrading the model’s performance.
